---
title: "Shiny Modules: From One App to Reusable Pieces"
author: "Dror Berel"
format:
  html:
    page-layout: full
    toc: true
    toc-location: left
    toc-depth: 3
    self-contained: true
execute:
  echo: true
editor_options: 
  chunk_output_type: console
editor: 
  markdown: 
    wrap: 72
---

# Learning goals

By the end of this lesson, you should be able to:

- explain what a Shiny module is
- recognize when regular Shiny code becomes repetitive
- turn a repeated UI + server pattern into a module
- reuse the same module more than once in one app

# What problem are modules solving?

In a small Shiny app, it is common to write everything in one place:

- all UI controls in one `fluidPage()`
- all server logic in one `server <- function(input, output, session)`

That is fine for small apps.

As the app grows, a few problems usually appear:

- repeated UI patterns
- repeated server logic
- harder-to-read code
- input and output names that start colliding

Modules help by packaging a piece of UI and server logic together so it
can be reused. A shiny module is a way to bind together UI and server
logic into a reusable component, as a function. This allows you to
create self-contained pieces of functionality that can be easily reused
across different parts of your app or even across different apps. The
binding is done through namespacing, which ensures that the inputs and
outputs of each module instance do not conflict with each other. In
practice, it means you add to the UI part: `ns <- NS(id)`

Then, instead of calling

`selectInput("dataset", "Choose a dataset", choices = c("mtcars", "iris"))`

you call:

`selectInput(ns("dataset"), "Choose a dataset", choices = c("mtcars", "iris"))`

In other words, a module allows you to create a mini Shiny app within
your main app, with its own UI and server logic, that can be
instantiated multiple times without worrying about naming conflicts.

# Step 1: Traditional approach, no modules

Here is a simple app with one dataset selector and one numeric summary.

```{r}
library(shiny)

ui <- fluidPage(
  titlePanel("No modules"),
  selectInput("dataset", "Choose a dataset", choices = c("mtcars", "iris")),
  selectInput("variable", "Choose a numeric variable", choices = NULL),
  verbatimTextOutput("summary")
)

server <- function(input, output, session) {
  data_reactive <- reactive({
    get(input$dataset)
  })

  observeEvent(data_reactive(), {
    numeric_vars <- names(data_reactive())[sapply(data_reactive(), is.numeric)]
    updateSelectInput(session, "variable", choices = numeric_vars, selected = numeric_vars[1])
  }, ignoreInit = FALSE)

  output$summary <- renderPrint({
    req(input$variable)
    summary(data_reactive()[[input$variable]])
  })
}

shinyApp(ui, server)
```

This works well.

But what if we want to show the same pattern twice in the same app?

- select a dataset
- select a numeric variable
- print a summary

Without modules, we usually copy and paste.

# Step 2: Repeating the traditional approach

This next app has the same feature twice.

```{r}
library(shiny)

ui <- fluidPage(
  titlePanel("Repeated pattern without modules"),
  fluidRow(
    column(
      width = 6,
      h3("Panel A"),
      selectInput("dataset_a", "Choose a dataset", choices = c("mtcars", "iris")),
      selectInput("variable_a", "Choose a numeric variable", choices = NULL),
      verbatimTextOutput("summary_a")
    ),
    column(
      width = 6,
      h3("Panel B"),
      selectInput("dataset_b", "Choose a dataset", choices = c("mtcars", "iris")),
      selectInput("variable_b", "Choose a numeric variable", choices = NULL),
      verbatimTextOutput("summary_b")
    )
  )
)

server <- function(input, output, session) {
  data_a <- reactive({
    get(input$dataset_a)
  })

  observeEvent(data_a(), {
    numeric_vars <- names(data_a())[sapply(data_a(), is.numeric)]
    updateSelectInput(session, "variable_a", choices = numeric_vars, selected = numeric_vars[1])
  }, ignoreInit = FALSE)

  output$summary_a <- renderPrint({
    req(input$variable_a)
    summary(data_a()[[input$variable_a]])
  })

  data_b <- reactive({
    get(input$dataset_b)
  })

  observeEvent(data_b(), {
    numeric_vars <- names(data_b())[sapply(data_b(), is.numeric)]
    updateSelectInput(session, "variable_b", choices = numeric_vars, selected = numeric_vars[1])
  }, ignoreInit = FALSE)

  output$summary_b <- renderPrint({
    req(input$variable_b)
    summary(data_b()[[input$variable_b]])
  })
}

shinyApp(ui, server)
```

This is still valid Shiny code, but now we can already see the problem:

- the structure is duplicated
- naming becomes awkward: `_a`, `_b`, `_c`, ...
- every future change has to be made in multiple places

# Step 3: Turn the repeated pattern into a module

A Shiny module usually has two parts:

- a UI function
- a server function

The UI function uses `NS()` to namespace input and output IDs.

The server function uses `moduleServer()` to create a private server
scope for that module instance.

## Module version

```{r}
library(shiny)

summaryModuleUI <- function(id, title = "Summary panel") {
  ns <- NS(id)

  tagList(
    h3(title),
    selectInput(ns("dataset"), "Choose a dataset", choices = c("mtcars", "iris")),
    selectInput(ns("variable"), "Choose a numeric variable", choices = NULL),
    verbatimTextOutput(ns("summary"))
  )
}

summaryModuleServer <- function(id) {
  moduleServer(id, function(input, output, session) {
    data_reactive <- reactive({
      get(input$dataset)
    })

    observeEvent(data_reactive(), {
      numeric_vars <- names(data_reactive())[sapply(data_reactive(), is.numeric)]
      updateSelectInput(session, "variable", choices = numeric_vars, selected = numeric_vars[1])
    }, ignoreInit = FALSE)

    output$summary <- renderPrint({
      req(input$variable)
      summary(data_reactive()[[input$variable]])
    })
  })
}

ui <- fluidPage(
  titlePanel("Using one module twice"),
  fluidRow(
    column(width = 6, summaryModuleUI("left_panel", "Panel A")),
    column(width = 6, summaryModuleUI("right_panel", "Panel B"))
  )
)

server <- function(input, output, session) {
  summaryModuleServer("left_panel")
  summaryModuleServer("right_panel")
}

shinyApp(ui, server)
```

# Why this is better

Now the repeated feature is defined once and reused twice.

Benefits:

- less copy-paste
- cleaner naming
- easier maintenance
- easier testing
- easier to build larger dashboards from smaller parts

The important idea is that each module instance gets its own namespace.

That means these are different internally:

- `left_panel-dataset`
- `right_panel-dataset`

You write one module, but Shiny keeps each instance separate.

# Short proof-of-concept examples

These are intentionally small.

## Proof of concept 1: A tiny text module

This example shows the basic shape of a reusable module.

```{r}
library(shiny)

textEchoUI <- function(id) {
  ns <- NS(id)

  tagList(
    textInput(ns("txt"), "Type something"),
    textOutput(ns("out"))
  )
}

textEchoServer <- function(id) {
  moduleServer(id, function(input, output, session) {
    output$out <- renderText({
      paste("You typed:", input$txt)
    })
  })
}

ui <- fluidPage(
  textEchoUI("one"),
  hr(),
  textEchoUI("two")
)

server <- function(input, output, session) {
  textEchoServer("one")
  textEchoServer("two")
}

shinyApp(ui, server)
```

## Proof of concept 2: A reusable plot module

This one is closer to a dashboard building block.

```{r}
library(shiny)

histogramUI <- function(id, title = "Histogram") {
  ns <- NS(id)

  tagList(
    h4(title),
    sliderInput(ns("bins"), "Bins", min = 5, max = 50, value = 15),
    plotOutput(ns("plot"))
  )
}

histogramServer <- function(id, data, var) {
  moduleServer(id, function(input, output, session) {
    output$plot <- renderPlot({
      hist(data()[[var]], breaks = input$bins, col = "steelblue", border = "white")
    })
  })
}

ui <- fluidPage(
  fluidRow(
    column(width = 6, histogramUI("mpg_plot", "Miles per gallon")),
    column(width = 6, histogramUI("hp_plot", "Horsepower"))
  )
)

server <- function(input, output, session) {
  data_reactive <- reactive(mtcars)

  histogramServer("mpg_plot", data = data_reactive, var = "mpg")
  histogramServer("hp_plot", data = data_reactive, var = "hp")
}

shinyApp(ui, server)
```

# A dashboard mindset

In dashboards, modules are especially useful because dashboards often
repeat patterns such as:

- filter panels
- value boxes
- plots with the same controls
- tables with the same formatting
- cards with similar layout and behavior

Instead of treating the dashboard as one big app, you can think of it as
a set of reusable building blocks.

# Practical rule of thumb

You probably do not need a module when:

- the app is very small
- the logic appears only once
- the code is still easy to read

You should consider a module when:

- the same pattern appears more than once
- a section of the app feels like a self-contained component
- naming conflicts are becoming annoying
- you want to reuse a feature across multiple apps

# One more small improvement

Modules can also return values.

For example, a filter module could return:

- a selected date range
- a selected category
- a filtered dataset

That makes modules useful not just for display, but also for
communication between parts of a larger app.

# Exercise ideas

Try these small changes:

1.  Add `faithful` as another dataset option.
2.  Modify the module so it shows both `summary()` and a histogram.
3.  Reuse the histogram module three times instead of two.
4.  Create a module for a shared sidebar filter.

# Key takeaway

Modules are not magic.

They are just a way to package:

- UI
- server logic
- namespacing

That makes repeated Shiny patterns easier to reuse and maintain, which
is especially helpful when building dashboards.


## Additional Resources
<https://github.com/drorberel/ComplexHeatmapTeal> 