---
title: "Write my own html widget"
author: "Dror Berel"
format:
  html:
    page-layout: full
    toc: true
    toc-location: left
    toc-depth: 3
    self-contained: true
execute:
  echo: true
editor_options: 
  chunk_output_type: console
editor: 
  markdown: 
    wrap: 72
---

## Goal

This lesson builds a very small custom `htmlwidget`.

The widget is intentionally simple: it draws horizontal bars from a data
frame. The important part is the shape of the code:

- an R function prepares data for the browser
- a JavaScript binding renders the widget
- Shiny calls the same R function with `rendermy_widget()`
- Crosstalk filters can hide and show rows in the widget



## Introduction
You may have already used htmlwidgets in R, such as `leaflet`, `plotly`, or `DT`. These widgets are built using a combination of R and JavaScript. In this lesson, we will create a simple custom htmlwidget that displays horizontal bars based on a data frame.

There are three main components to creating a custom htmlwidget:
the JS binding, the R function, and the Shiny integration. The JS binding is responsible for rendering the widget in the browser, while the R function prepares the data for the browser. The Shiny integration allows us to use the widget in a Shiny app.

Specifically, we will need these functions:
xx.JS
my_widget_Output() (the euivalent of DT::DTOutput())
render_my_widget() (the equivalent of DT::renderDT())


## Packages

Install these packages if needed:

```{r}
# install.packages(c("htmlwidgets", "htmltools", "shiny", "crosstalk", "DT"))
```

Load the packages:

```{r}
library(htmlwidgets)
library(htmltools)
library(shiny)
library(crosstalk)
library(DT)
```

## The Widget Code

In a real package, the JavaScript below would usually live in a separate
file under `inst/htmlwidgets/`. For a proof of concept, an inline
`<script>` tag keeps everything in one lesson file.

```{r}
my_widget_JS <- function() {
  tags$script(HTML(
    "
(function registermy_widgetBinding() {
  if (!window.HTMLWidgets) {
    setTimeout(registermy_widgetBinding, 20);
    return;
  }

  if (window.my_widgetBindingRegistered) {
    return;
  }

  window.my_widgetBindingRegistered = true;

HTMLWidgets.widget({
  name: 'my_widget',
  type: 'output',

  factory: function(el, width, height) {
    var currentX = null;
    var filterHandle = null;
    var filterCallback = null;
    var activeKeys = null;

    function draw(x) {
      currentX = x;
      var rows = x.rows || [];

      if (activeKeys !== null) {
        var keep = new Set(activeKeys.map(String));
        rows = rows.filter(function(row) {
          return keep.has(String(row.key));
        });
      }

      var maxValue = Math.max.apply(
        null,
        rows.map(function(row) { return row.value; }).concat([1])
      );

      el.innerHTML = '';
      el.style.fontFamily = 'system-ui, -apple-system, BlinkMacSystemFont, Segoe UI, sans-serif';
      el.style.boxSizing = 'border-box';
      el.style.padding = '8px';

      if (rows.length === 0) {
        var empty = document.createElement('div');
        empty.textContent = 'No rows match the current filter.';
        empty.style.color = '#666';
        el.appendChild(empty);
        return;
      }

      rows.forEach(function(row) {
        var line = document.createElement('div');
        line.style.display = 'grid';
        line.style.gridTemplateColumns = '130px 1fr 48px';
        line.style.gap = '8px';
        line.style.alignItems = 'center';
        line.style.margin = '8px 0';

        var label = document.createElement('div');
        label.textContent = row.label;
        label.style.overflow = 'hidden';
        label.style.textOverflow = 'ellipsis';
        label.style.whiteSpace = 'nowrap';

        var track = document.createElement('div');
        track.style.background = '#e8edf3';
        track.style.height = '18px';
        track.style.borderRadius = '4px';

        var bar = document.createElement('div');
        bar.style.width = ((row.value / maxValue) * 100) + '%';
        bar.style.height = '18px';
        bar.style.borderRadius = '4px';
        bar.style.background = '#2a6f9e';
        track.appendChild(bar);

        var value = document.createElement('div');
        value.textContent = row.value;
        value.style.textAlign = 'right';
        value.style.fontVariantNumeric = 'tabular-nums';

        line.appendChild(label);
        line.appendChild(track);
        line.appendChild(value);
        el.appendChild(line);
      });
    }

    return {
      renderValue: function(x) {
        if (filterHandle && filterCallback && filterHandle.off) {
          filterHandle.off('change', filterCallback);
          filterHandle = null;
          filterCallback = null;
        }

        activeKeys = null;

        if (x.group && window.crosstalk) {
          filterHandle = new crosstalk.FilterHandle(x.group);
          filterCallback = function(event) {
            activeKeys = event.value === null ? null : event.value;
            draw(currentX);
          };
          filterHandle.on('change', filterCallback);
        }

        draw(x);
      },

      resize: function(width, height) {
        if (currentX) {
          draw(currentX);
        }
      }
    };
  }
});
})();
    "
  ))
}

my_widget_bind <- function(data, label, value, width = "100%", height = 260) {
  label_name <- deparse(substitute(label))
  value_name <- deparse(substitute(value))

  if (crosstalk::is.SharedData(data)) {
    raw_data <- data$data(withSelection = FALSE, withFilter = FALSE)
    keys <- data$key()
    group <- data$groupName()
  } else {
    raw_data <- data
    keys <- seq_len(nrow(raw_data))
    group <- NULL
  }

  rows <- lapply(seq_len(nrow(raw_data)), function(i) {
    list(
      key = as.character(keys[i]),
      label = as.character(raw_data[[label_name]][i]),
      value = as.numeric(raw_data[[value_name]][i])
    )
  })

  htmlwidgets::createWidget(
    name = "my_widget",
    x = list(rows = rows, group = group),
    width = width,
    height = height,
    package = "htmlwidgets"
  )
}

my_widgetOutput <- function(outputId, width = "100%", height = "260px") {
  htmlwidgets::shinyWidgetOutput(
    outputId = outputId,
    name = "my_widget",
    width = width,
    height = height,
    package = "htmlwidgets"
  )
}

rendermy_widget <- function(expr, env = parent.frame(), quoted = FALSE) {
  if (!quoted) {
    expr <- substitute(expr)
  }

  htmlwidgets::shinyRenderWidget(
    expr = expr,
    outputFunction = my_widgetOutput,
    env = env,
    quoted = TRUE
  )
}
```

## Plain R Call

The widget function accepts a data frame and the names of the label and
value columns.

```{r}
fruit <- data.frame(
  fruit = c("Apple", "Banana", "Blueberry", "Orange", "Pear"),
  family = c("Pome", "Berry", "Berry", "Citrus", "Pome"),
  value = c(10, 16, 7, 12, 5)
)

tagList(
  my_widget_JS(),
  my_widget_bind(fruit, fruit, value)
)
```

## Shiny Call (without crosstalk)

In Shiny, the pattern is the same as other htmlwidgets:

- put the JavaScript binding in the UI once
- place `my_widgetOutput()` where the widget should appear
- return `my_widget_bind()` from `rendermy_widget()`

This chunk is marked `eval=FALSE`, so the rendered Quarto document shows
the app source code instead of launching a Shiny app while rendering. Run
it interactively after running the widget-definition chunk above.

```{r, eval=FALSE}
library(shiny)
library(htmlwidgets)
library(htmltools)

fruit <- data.frame(
  fruit = c("Apple", "Banana", "Blueberry", "Orange", "Pear"),
  family = c("Pome", "Berry", "Berry", "Citrus", "Pome"),
  value = c(10, 16, 7, 12, 5)
)

ui <- fluidPage(
  my_widget_JS(),
  titlePanel("Custom htmlwidget"),
  sliderInput("scale", "Scale values", min = 1, max = 5, value = 1),
  my_widgetOutput("bars", height = "280px")
)

server <- function(input, output, session) {
  output$bars <- rendermy_widget({
    scaled_fruit <- fruit
    scaled_fruit$value[scaled_fruit$fruit == "Banana"] <-
      scaled_fruit$value[scaled_fruit$fruit == "Banana"] * input$scale

    my_widget_bind(scaled_fruit, fruit, value)
  })
}

shinyApp(ui, server)

```

## Crosstalk Call

To support Crosstalk, the widget needs two pieces of metadata:

- `key`: the row identifier
- `group`: the Crosstalk group shared by filters and widgets

The R function above detects `SharedData` and sends both values to
JavaScript. The JavaScript binding listens to
`crosstalk.FilterHandle(group)` and redraws the bars when filters change.

This example works in a regular rendered HTML document.

```{r}
shared_fruit <- SharedData$new(
  data = fruit,
  key = ~fruit,
  group = "fruit"
)

tagList(
  my_widget_JS(),
  bscols(
    widths = c(3, 9),
    list(
      filter_select(
        id = "family_filter",
        label = "Fruit family",
        sharedData = shared_fruit,
        group = ~family,
        multiple = TRUE
      ),
      filter_slider(
        id = "value_filter",
        label = "Value",
        sharedData = shared_fruit,
        column = ~value,
        step = 1
      )
    ),
    my_widget_bind(shared_fruit, fruit, value)
  )
)
```

## Shiny + Crosstalk Call

This final example uses all three parts together:

- Shiny owns the page and can add server-side reactivity
- Crosstalk filters run in the browser
- the custom htmlwidget listens to the Crosstalk filter state

Without the crosstalk filters in the left column, the bars and table in the right column would not respond to user input. With the filters, users can hide and show rows in both the table and the custom widget at the same time.

Save this as `app.R`, or run it from an R session after defining
`my_widget_JS()`, `my_widget_bind()`, `my_widgetOutput()`, and
`rendermy_widget()`.

```{r, eval=FALSE}
library(shiny)
library(htmlwidgets)
library(htmltools)
library(crosstalk)
library(DT)

fruit <- data.frame(
  fruit = c("Apple", "Banana", "Blueberry", "Orange", "Pear"),
  family = c("Pome", "Berry", "Berry", "Citrus", "Pome"),
  value = c(10, 16, 7, 12, 5)
)

shared_fruit <- SharedData$new(
  data = fruit,
  key = ~fruit,
  group = "fruit"
)

corsstalk_ui <- fluidPage(
  my_widget_JS(),
  titlePanel("Shiny + Crosstalk + custom htmlwidget"),
  fluidRow(
    column(
      width = 3,
      filter_select(
        id = "family_filter_app",
        label = "Fruit family",
        sharedData = shared_fruit,
        group = ~family,
        multiple = TRUE
      ),
      filter_slider(
        id = "value_filter_app",
        label = "Value",
        sharedData = shared_fruit,
        column = ~value,
        step = 1
      )
    ),
    column(
      width = 9,
      my_widgetOutput("bars", height = "280px"),
      DTOutput("table")
    )
  )
)

corsstalk_server <- function(input, output, session) {
  output$bars <- rendermy_widget({
    my_widget_bind(shared_fruit, fruit, value)
  })

  output$table <- renderDT({
    datatable(shared_fruit, rownames = FALSE)
  })
}

shinyApp(corsstalk_ui, corsstalk_server)

```

## What To Notice

The widget does not know anything about Shiny. Shiny only needs the
usual htmlwidget output and render functions.

Crosstalk is different: the widget must explicitly opt in by receiving a
`group` and listening for browser-side filter events. Without that small
JavaScript listener, Crosstalk filters would not affect the custom
widget.
