---
title: "write my own html widget"
author: "Dror Berel"
format:
  html:
    toc: true
    code-fold: false
    self-contained: true
execute:
  echo: true
editor_options: 
  chunk_output_type: console
editor: 
  markdown: 
    wrap: 72
---

## Goal

This lesson builds a very small custom `htmlwidget`.

The widget is intentionally simple: it draws horizontal bars from a data
frame. The important part is the shape of the code:

- an R function prepares data for the browser
- a JavaScript binding renders the widget
- Shiny calls the same R function with `renderMiniBars()`
- Crosstalk filters can hide and show rows in the widget

## Packages

Install these packages if needed:

```{r}
# install.packages(c("htmlwidgets", "htmltools", "shiny", "crosstalk", "DT"))
```

Load the packages:

```{r}
library(htmlwidgets)
library(htmltools)
library(shiny)
library(crosstalk)
library(DT)
```

## The Widget Code

In a real package, the JavaScript below would usually live in a separate
file under `inst/htmlwidgets/`. For a proof of concept, an inline
`<script>` tag keeps everything in one lesson file.

```{r}
mini_bars_binding <- function() {
  tags$script(HTML(
    "
HTMLWidgets.widget({
  name: 'miniBars',
  type: 'output',

  factory: function(el, width, height) {
    var currentX = null;
    var filterHandle = null;
    var filterCallback = null;
    var activeKeys = null;

    function draw(x) {
      currentX = x;
      var rows = x.rows || [];

      if (activeKeys !== null) {
        var keep = new Set(activeKeys.map(String));
        rows = rows.filter(function(row) {
          return keep.has(String(row.key));
        });
      }

      var maxValue = Math.max.apply(
        null,
        rows.map(function(row) { return row.value; }).concat([1])
      );

      el.innerHTML = '';
      el.style.fontFamily = 'system-ui, -apple-system, BlinkMacSystemFont, Segoe UI, sans-serif';
      el.style.boxSizing = 'border-box';
      el.style.padding = '8px';

      if (rows.length === 0) {
        var empty = document.createElement('div');
        empty.textContent = 'No rows match the current filter.';
        empty.style.color = '#666';
        el.appendChild(empty);
        return;
      }

      rows.forEach(function(row) {
        var line = document.createElement('div');
        line.style.display = 'grid';
        line.style.gridTemplateColumns = '130px 1fr 48px';
        line.style.gap = '8px';
        line.style.alignItems = 'center';
        line.style.margin = '8px 0';

        var label = document.createElement('div');
        label.textContent = row.label;
        label.style.overflow = 'hidden';
        label.style.textOverflow = 'ellipsis';
        label.style.whiteSpace = 'nowrap';

        var track = document.createElement('div');
        track.style.background = '#e8edf3';
        track.style.height = '18px';
        track.style.borderRadius = '4px';

        var bar = document.createElement('div');
        bar.style.width = ((row.value / maxValue) * 100) + '%';
        bar.style.height = '18px';
        bar.style.borderRadius = '4px';
        bar.style.background = '#2a6f9e';
        track.appendChild(bar);

        var value = document.createElement('div');
        value.textContent = row.value;
        value.style.textAlign = 'right';
        value.style.fontVariantNumeric = 'tabular-nums';

        line.appendChild(label);
        line.appendChild(track);
        line.appendChild(value);
        el.appendChild(line);
      });
    }

    return {
      renderValue: function(x) {
        if (filterHandle && filterCallback && filterHandle.off) {
          filterHandle.off('change', filterCallback);
          filterHandle = null;
          filterCallback = null;
        }

        activeKeys = null;

        if (x.group && window.crosstalk) {
          filterHandle = new crosstalk.FilterHandle(x.group);
          filterCallback = function(event) {
            activeKeys = event.value === null ? null : event.value;
            draw(currentX);
          };
          filterHandle.on('change', filterCallback);
        }

        draw(x);
      },

      resize: function(width, height) {
        if (currentX) {
          draw(currentX);
        }
      }
    };
  }
});
    "
  ))
}

mini_bars <- function(data, label, value, width = "100%", height = 260) {
  label_name <- deparse(substitute(label))
  value_name <- deparse(substitute(value))

  if (crosstalk::is.SharedData(data)) {
    raw_data <- data$data(withSelection = FALSE, withFilter = FALSE)
    keys <- data$key()
    group <- data$groupName()
  } else {
    raw_data <- data
    keys <- seq_len(nrow(raw_data))
    group <- NULL
  }

  rows <- data.frame(
    key = as.character(keys),
    label = as.character(raw_data[[label_name]]),
    value = as.numeric(raw_data[[value_name]]),
    stringsAsFactors = FALSE
  )

  htmlwidgets::createWidget(
    name = "miniBars",
    x = list(rows = rows, group = group),
    width = width,
    height = height,
    package = "htmlwidgets"
  )
}

miniBarsOutput <- function(outputId, width = "100%", height = "260px") {
  htmlwidgets::shinyWidgetOutput(
    outputId = outputId,
    name = "miniBars",
    width = width,
    height = height,
    package = "htmlwidgets"
  )
}

renderMiniBars <- function(expr, env = parent.frame(), quoted = FALSE) {
  if (!quoted) {
    expr <- substitute(expr)
  }

  htmlwidgets::shinyRenderWidget(
    expr = expr,
    outputFunction = miniBarsOutput,
    env = env,
    quoted = TRUE
  )
}
```

## Plain R Call

The widget function accepts a data frame and the names of the label and
value columns.

```{r}
fruit <- data.frame(
  fruit = c("Apple", "Banana", "Blueberry", "Orange", "Pear"),
  family = c("Pome", "Berry", "Berry", "Citrus", "Pome"),
  value = c(10, 16, 7, 12, 5)
)

tagList(
  mini_bars_binding(),
  mini_bars(fruit, fruit, value)
)
```

## Shiny Call

In Shiny, the pattern is the same as other htmlwidgets:

- put the JavaScript binding in the UI once
- place `miniBarsOutput()` where the widget should appear
- return `mini_bars()` from `renderMiniBars()`

```{r, eval=FALSE}
library(shiny)
library(htmlwidgets)
library(htmltools)

fruit <- data.frame(
  fruit = c("Apple", "Banana", "Blueberry", "Orange", "Pear"),
  family = c("Pome", "Berry", "Berry", "Citrus", "Pome"),
  value = c(10, 16, 7, 12, 5)
)

ui <- fluidPage(
  mini_bars_binding(),
  titlePanel("Custom htmlwidget"),
  sliderInput("scale", "Scale values", min = 1, max = 5, value = 1),
  miniBarsOutput("bars", height = "280px")
)

server <- function(input, output, session) {
  output$bars <- renderMiniBars({
    scaled_fruit <- fruit
    scaled_fruit$value <- scaled_fruit$value * input$scale

    mini_bars(scaled_fruit, fruit, value)
  })
}

shinyApp(ui, server)
```

## Crosstalk Call

To support Crosstalk, the widget needs two pieces of metadata:

- `key`: the row identifier
- `group`: the Crosstalk group shared by filters and widgets

The R function above detects `SharedData` and sends both values to
JavaScript. The JavaScript binding listens to
`crosstalk.FilterHandle(group)` and redraws the bars when filters change.

This example works in a regular rendered HTML document.

```{r}
shared_fruit <- SharedData$new(
  data = fruit,
  key = ~fruit,
  group = "fruit"
)

tagList(
  mini_bars_binding(),
  bscols(
    widths = c(3, 9),
    list(
      filter_select(
        id = "family_filter",
        label = "Fruit family",
        sharedData = shared_fruit,
        group = ~family,
        multiple = TRUE
      ),
      filter_slider(
        id = "value_filter",
        label = "Value",
        sharedData = shared_fruit,
        column = ~value,
        step = 1
      )
    ),
    mini_bars(shared_fruit, fruit, value)
  )
)
```

## Shiny + Crosstalk Call

This final example uses all three parts together:

- Shiny owns the page and can add server-side reactivity
- Crosstalk filters run in the browser
- the custom htmlwidget listens to the Crosstalk filter state

Save this as `app.R`, or run it from an R session after defining
`mini_bars_binding()`, `mini_bars()`, `miniBarsOutput()`, and
`renderMiniBars()`.

```{r, eval=FALSE}
library(shiny)
library(htmlwidgets)
library(htmltools)
library(crosstalk)
library(DT)

fruit <- data.frame(
  fruit = c("Apple", "Banana", "Blueberry", "Orange", "Pear"),
  family = c("Pome", "Berry", "Berry", "Citrus", "Pome"),
  value = c(10, 16, 7, 12, 5)
)

shared_fruit <- SharedData$new(
  data = fruit,
  key = ~fruit,
  group = "fruit"
)

ui <- fluidPage(
  mini_bars_binding(),
  titlePanel("Shiny + Crosstalk + custom htmlwidget"),
  fluidRow(
    column(
      width = 3,
      filter_select(
        id = "family_filter_app",
        label = "Fruit family",
        sharedData = shared_fruit,
        group = ~family,
        multiple = TRUE
      ),
      filter_slider(
        id = "value_filter_app",
        label = "Value",
        sharedData = shared_fruit,
        column = ~value,
        step = 1
      )
    ),
    column(
      width = 9,
      miniBarsOutput("bars", height = "280px"),
      DTOutput("table")
    )
  )
)

server <- function(input, output, session) {
  output$bars <- renderMiniBars({
    mini_bars(shared_fruit, fruit, value)
  })

  output$table <- renderDT({
    datatable(shared_fruit, rownames = FALSE)
  })
}

shinyApp(ui, server)
```

## What To Notice

The widget does not know anything about Shiny. Shiny only needs the
usual htmlwidget output and render functions.

Crosstalk is different: the widget must explicitly opt in by receiving a
`group` and listening for browser-side filter events. Without that small
JavaScript listener, Crosstalk filters would not affect the custom
widget.
