---
title: "bslib: Multi-Page Shiny Apps with Shared Sidebar Controls"
author: "Dror Berel"
format:
  html:
    page-layout: full
    toc: true
    toc-location: left
    toc-depth: 3
    self-contained: true
execute:
  echo: true
editor_options: 
  chunk_output_type: console
editor: 
  markdown: 
    wrap: 72
---

# Learning goals

By the end of this lesson, you should be able to:

- use `bslib` as the main layout system for a Shiny app
- create a Shiny app with multiple tabs or pages
- place project-level inputs in one sidebar
- reuse the same sidebar inputs across all tabs
- organize outputs with `card()`, `layout_columns()`, and value boxes

# Why `bslib`?

Base Shiny layouts such as `fluidPage()`, `sidebarLayout()`, and
`tabsetPanel()` are still useful, but many modern Shiny apps need more
structure:

- a consistent page layout
- multiple views of the same project
- cards that group related outputs
- sidebars that can be opened, closed, and shared across pages
- Bootstrap styling without writing much CSS

The `bslib` package gives Shiny a more modern layout vocabulary. In
this tutorial, we will build a small app where one sidebar controls all
tabs.

# Packages

Install these packages if needed:

```{r}
# install.packages(c("shiny", "bslib", "bsicons", "dplyr", "ggplot2", "DT"))
```

Load the packages:

```{r}
library(shiny)
library(bslib)
library(dplyr)
library(ggplot2)
library(DT)
```

# App idea

We will build a small project dashboard using `mtcars`.

The app has one project-level sidebar with controls for:

- cylinder groups
- transmission type
- the main numeric metric
- whether to show a trend line

Those inputs are not local to one tab. They apply to every tab:

- the overview tab shows summary values
- the plot tab shows a filtered scatter plot
- the table tab shows the filtered rows
- the notes tab explains the current project state

# Prepare the data

The app uses `mtcars`, but we first create friendlier labels.

```{r}
cars <- mtcars |>
  tibble::rownames_to_column("car") |>
  mutate(
    cyl = factor(
      cyl,
      levels = c(4, 6, 8),
      labels = c("4 cylinders", "6 cylinders", "8 cylinders")
    ),
    am = factor(
      am,
      levels = c(0, 1),
      labels = c("Automatic", "Manual")
    ),
    gear = factor(gear),
    carb = factor(carb)
  )

head(cars)
```

# Step 1: Create the page shell

The outer layout is `page_navbar()`. Each `nav_panel()` becomes one page
inside the app.

```{r, eval=FALSE}
ui_panels_only <- page_navbar(
  title = "Motor Trend Explorer",
  theme = bs_theme(bootswatch = "flatly"),
  nav_panel("Overview"),
  nav_panel("Plot"),
  nav_panel("Table"),
  nav_panel("Notes")
)
```

This gives us pages, but not yet a shared sidebar.

# Step 2: Add a project-level sidebar

The important pattern is that the sidebar belongs to `page_navbar()`,
not to a single tab. Because of that, the inputs are available to every
tab in the app.

```{r, eval=FALSE}
ui_with_sidebar <- page_navbar(
  title = "Motor Trend Explorer",
  theme = bs_theme(bootswatch = "flatly"),
  sidebar = sidebar(
    title = "Project controls",
    checkboxGroupInput(
      inputId = "cylinders",
      label = "Cylinders",
      choices = levels(cars$cyl),
      selected = levels(cars$cyl)
    ),
    checkboxGroupInput(
      inputId = "transmission",
      label = "Transmission",
      choices = levels(cars$am),
      selected = levels(cars$am)
    ),
    selectInput(
      inputId = "metric",
      label = "Main metric",
      choices = c(
        "Miles per gallon" = "mpg",
        "Horsepower" = "hp",
        "Weight" = "wt",
        "Quarter mile time" = "qsec"
      ),
      selected = "mpg"
    ),
    checkboxInput(
      inputId = "show_smooth",
      label = "Show trend line",
      value = TRUE
    )
  ),
  nav_panel("Overview"),
  nav_panel("Plot"),
  nav_panel("Table"),
  nav_panel("Notes")
)
```

# Step 3: Add cards and outputs

`bslib::card()` is useful for grouping related content. Value boxes are
good for high-level numbers. `layout_columns()` creates responsive
columns that adjust as the screen changes.

```{r, eval=FALSE}
nav_panel(
  "Overview",
  layout_columns(
    value_box(
      title = "Cars",
      value = textOutput("n_cars"),
      showcase = bsicons::bs_icon("car-front")
    ),
    value_box(
      title = "Average metric",
      value = textOutput("avg_metric"),
      showcase = bsicons::bs_icon("speedometer2")
    ),
    value_box(
      title = "Best car",
      value = textOutput("best_car"),
      showcase = bsicons::bs_icon("trophy")
    )
  ),
  card(
    card_header("Summary by cylinder group"),
    tableOutput("cylinder_summary")
  )
)
```

In a real app, each tab can have a different purpose while still using
the same project-level filters.

# Step 4: Define shared server logic

The key server pattern is to create one filtered reactive expression.
All tabs use this same reactive object.

```{r, eval=FALSE}
filtered_cars <- reactive({
  req(input$cylinders, input$transmission)

  cars |>
    filter(
      cyl %in% input$cylinders,
      am %in% input$transmission
    )
})
```

This is the core idea of project-level UI parameters:

- define the inputs once
- define the filtering logic once
- reuse the resulting reactive data everywhere

# Complete app

Save the following code as `app.R`, or run it from an R session.

```{r, eval=FALSE}
library(shiny)
library(bslib)
library(dplyr)
library(ggplot2)
library(DT)

cars <- mtcars |>
  tibble::rownames_to_column("car") |>
  mutate(
    cyl = factor(
      cyl,
      levels = c(4, 6, 8),
      labels = c("4 cylinders", "6 cylinders", "8 cylinders")
    ),
    am = factor(
      am,
      levels = c(0, 1),
      labels = c("Automatic", "Manual")
    ),
    gear = factor(gear),
    carb = factor(carb)
  )

metric_choices <- c(
  "Miles per gallon" = "mpg",
  "Horsepower" = "hp",
  "Weight" = "wt",
  "Quarter mile time" = "qsec"
)

ui <- page_navbar(
  title = "Motor Trend Explorer",
  theme = bs_theme(
    version = 5,
    bootswatch = "flatly",
    primary = "#2c7fb8"
  ),
  sidebar = sidebar(
    title = "Project controls",
    width = 300,
    checkboxGroupInput(
      inputId = "cylinders",
      label = "Cylinders",
      choices = levels(cars$cyl),
      selected = levels(cars$cyl)
    ),
    checkboxGroupInput(
      inputId = "transmission",
      label = "Transmission",
      choices = levels(cars$am),
      selected = levels(cars$am)
    ),
    selectInput(
      inputId = "metric",
      label = "Main metric",
      choices = metric_choices,
      selected = "mpg"
    ),
    checkboxInput(
      inputId = "show_smooth",
      label = "Show trend line",
      value = TRUE
    )
  ),
  nav_panel(
    "Overview",
    layout_columns(
      value_box(
        title = "Cars",
        value = textOutput("n_cars"),
        showcase = bsicons::bs_icon("car-front")
      ),
      value_box(
        title = "Average metric",
        value = textOutput("avg_metric"),
        showcase = bsicons::bs_icon("speedometer2")
      ),
      value_box(
        title = "Best car",
        value = textOutput("best_car"),
        showcase = bsicons::bs_icon("trophy")
      )
    ),
    card(
      card_header("Summary by cylinder group"),
      tableOutput("cylinder_summary")
    )
  ),
  nav_panel(
    "Plot",
    card(
      full_screen = TRUE,
      card_header(textOutput("plot_title")),
      plotOutput("metric_plot", height = 480)
    )
  ),
  nav_panel(
    "Table",
    card(
      full_screen = TRUE,
      card_header("Filtered cars"),
      DTOutput("cars_table")
    )
  ),
  nav_panel(
    "Notes",
    layout_columns(
      card(
        card_header("Current project state"),
        verbatimTextOutput("project_state")
      ),
      card(
        card_header("Pattern"),
        tags$p(
          "The sidebar is attached to page_navbar(), so the controls are ",
          "available across all nav panels."
        ),
        tags$p(
          "The server creates one filtered reactive dataset and every tab ",
          "uses that same reactive object."
        )
      )
    )
  )
)

server <- function(input, output, session) {
  filtered_cars <- reactive({
    req(input$cylinders, input$transmission)

    cars |>
      filter(
        cyl %in% input$cylinders,
        am %in% input$transmission
      )
  })

  selected_metric_label <- reactive({
    names(metric_choices)[match(input$metric, metric_choices)]
  })

  output$n_cars <- renderText({
    nrow(filtered_cars())
  })

  output$avg_metric <- renderText({
    value <- mean(filtered_cars()[[input$metric]], na.rm = TRUE)
    round(value, 1)
  })

  output$best_car <- renderText({
    data <- filtered_cars()
    req(nrow(data) > 0)

    data$car[which.max(data[[input$metric]])]
  })

  output$cylinder_summary <- renderTable({
    filtered_cars() |>
      group_by(cyl) |>
      summarise(
        cars = n(),
        avg_metric = round(mean(.data[[input$metric]], na.rm = TRUE), 1),
        avg_mpg = round(mean(mpg, na.rm = TRUE), 1),
        .groups = "drop"
      )
  })

  output$plot_title <- renderText({
    paste(selected_metric_label(), "by vehicle weight")
  })

  output$metric_plot <- renderPlot({
    p <- ggplot(
      filtered_cars(),
      aes(x = wt, y = .data[[input$metric]], color = cyl)
    ) +
      geom_point(size = 3, alpha = 0.85) +
      labs(
        x = "Weight",
        y = selected_metric_label(),
        color = "Cylinders"
      ) +
      theme_minimal(base_size = 13)

    if (isTRUE(input$show_smooth)) {
      p <- p + geom_smooth(method = "lm", se = FALSE, linewidth = 0.8)
    }

    p
  })

  output$cars_table <- renderDT({
    filtered_cars() |>
      select(car, cyl, am, mpg, hp, wt, qsec, gear, carb) |>
      datatable(
        rownames = FALSE,
        options = list(pageLength = 10, scrollX = TRUE)
      )
  })

  output$project_state <- renderPrint({
    list(
      cylinders = input$cylinders,
      transmission = input$transmission,
      metric = selected_metric_label(),
      rows_after_filtering = nrow(filtered_cars())
    )
  })
}

shinyApp(ui, server)
```

# What to notice

## The sidebar is global

The sidebar is passed directly to `page_navbar()`:

```{r, eval=FALSE}
page_navbar(
  sidebar = sidebar(...),
  nav_panel("Overview", ...),
  nav_panel("Plot", ...),
  nav_panel("Table", ...)
)
```

That means the sidebar is part of the whole app layout. It is not owned
by one tab.

## The reactive data is shared

The server creates one filtered dataset:

```{r, eval=FALSE}
filtered_cars <- reactive({
  cars |>
    filter(
      cyl %in% input$cylinders,
      am %in% input$transmission
    )
})
```

Every output uses `filtered_cars()`. This keeps the behavior consistent
across tabs.

## The tabs are only views

The tabs do not each define their own filtering logic. They are
different views of the same project state:

- overview numbers
- plot
- table
- notes

This makes the app easier to maintain. When the project-level filters
change, all tabs update automatically.

# Moving a control into one tab

Not every input belongs in the project-level sidebar. A good rule is:

- put inputs in the sidebar when they affect the whole app
- put inputs inside a tab when they only affect that tab

In this app, `show_smooth` only affects the plot. It does not change the
overview values, the table, or the project notes. That makes it a good
candidate for moving into the Plot tab.

## Remove it from the sidebar

First, remove this input from the `sidebar()`:

```{r, eval=FALSE}
checkboxInput(
  inputId = "show_smooth",
  label = "Show trend line",
  value = TRUE
)
```

After removing it, the sidebar contains only project-level controls:

```{r, eval=FALSE}
sidebar = sidebar(
  title = "Project controls",
  width = 300,
  checkboxGroupInput(
    inputId = "cylinders",
    label = "Cylinders",
    choices = levels(cars$cyl),
    selected = levels(cars$cyl)
  ),
  checkboxGroupInput(
    inputId = "transmission",
    label = "Transmission",
    choices = levels(cars$am),
    selected = levels(cars$am)
  ),
  selectInput(
    inputId = "metric",
    label = "Main metric",
    choices = metric_choices,
    selected = "mpg"
  )
)
```

## Add it to the Plot tab

Then add the checkbox to the Plot tab. One simple option is to place it
inside the plot card, above the plot output:

```{r, eval=FALSE}
nav_panel(
  "Plot",
  card(
    full_screen = TRUE,
    card_header(textOutput("plot_title")),
    checkboxInput(
      inputId = "show_smooth",
      label = "Show trend line",
      value = TRUE
    ),
    plotOutput("metric_plot", height = 480)
  )
)
```

The server code does not need to change. This line still works:

```{r, eval=FALSE}
if (isTRUE(input$show_smooth)) {
  p <- p + geom_smooth(method = "lm", se = FALSE, linewidth = 0.8)
}
```

The input is in a different place in the UI, but it still has the same
`inputId`. In a regular Shiny app, input IDs belong to the app session,
not to a visual container. Moving the input from the sidebar into the
Plot tab changes where the user sees it, not how the server reads it.

## Why this is better

This version makes the app easier to understand:

- the sidebar controls the whole project state
- the Plot tab owns the plot-only option
- the server still uses the same `input$show_smooth` value

As apps grow, this distinction helps keep sidebars from becoming a
catch-all for every possible option.

# Case study: Use `page_navbar(id = "page")`

So far, the tabs are only visual containers. The server does not know
which tab the user is viewing.

Sometimes that is fine. But in larger apps, it is useful for the server
to know the active page. Add an `id` to `page_navbar()`:

```{r, eval=FALSE}
ui <- page_navbar(
  id = "page",
  title = "Motor Trend Explorer",
  theme = bs_theme(
    version = 5,
    bootswatch = "flatly",
    primary = "#2c7fb8"
  ),
  sidebar = sidebar(...),
  nav_panel("Overview", ...),
  nav_panel("Plot", ...),
  nav_panel("Table", ...),
  nav_panel("Notes", ...)
)
```

Now Shiny creates `input$page`. Its value is the currently selected
navigation panel.

## Display the active page

Add a small status line to the sidebar:

```{r, eval=FALSE}
sidebar = sidebar(
  title = "Project controls",
  width = 300,
  checkboxGroupInput(...),
  checkboxGroupInput(...),
  selectInput(...),
  checkboxInput(...),
  hr(),
  textOutput("active_page")
)
```

Then read `input$page` in the server:

```{r, eval=FALSE}
output$active_page <- renderText({
  paste("Current page:", input$page)
})
```

This is the simplest use case: the UI can show navigation state without
duplicating it manually.

## Make the sidebar page-aware

The more useful pattern is to change helper text or controls depending
on the active page.

Add this placeholder to the sidebar:

```{r, eval=FALSE}
uiOutput("page_hint")
```

Then render different text for each page:

```{r, eval=FALSE}
output$page_hint <- renderUI({
  switch(
    input$page,
    "Overview" = tags$p("Use this page to check high-level project metrics."),
    "Plot" = tags$p("Use this page to inspect relationships in the filtered data."),
    "Table" = tags$p("Use this page to review and export filtered records."),
    "Notes" = tags$p("Use this page to inspect the current app state."),
    tags$p("Choose a page from the navigation bar.")
  )
})
```

This keeps the sidebar global, but lets it respond to the user's current
workflow.

## Defer expensive work until a page is opened

Another practical use is avoiding expensive output work until a tab is
actually visible. For example, imagine the table is slow because it
comes from a database or a large file.

```{r, eval=FALSE}
output$cars_table <- renderDT({
  req(input$page == "Table")

  filtered_cars() |>
    select(car, cyl, am, mpg, hp, wt, qsec, gear, carb) |>
    datatable(
      rownames = FALSE,
      options = list(pageLength = 10, scrollX = TRUE)
    )
})
```

With this pattern, the table code only runs after the user navigates to
the Table page.

## Track app usage

You can also observe page changes. This is useful for lightweight
analytics, debugging, or saving user progress.

```{r, eval=FALSE}
observeEvent(input$page, {
  message("User opened page: ", input$page)
})
```

In a production app, the same pattern could write to a log table, update
a bookmark state, or trigger a page-specific data refresh.

## When `id = "page"` is helpful

Use `page_navbar(id = "page")` when the active tab should become part of
the app's reactive state.

Good cases include:

- showing page-specific instructions or validation messages
- deferring slow plots, tables, models, or API calls until needed
- tracking which pages users visit
- preserving or restoring the user's current page
- enabling page-specific downloads or actions
- warning users before they leave a page with unsaved selections
- changing sidebar controls based on the active workflow

Avoid using it when tabs are only static layout sections and the server
does not need to react to navigation.

# Exercise

Extend the app by adding one more project-level control to the sidebar.

Good options:

- minimum horsepower
- gear count
- number of rows shown in the table
- plot color variable

Then update at least two tabs so they respond to the new control.

# Key takeaway

Use a `bslib` page-level sidebar when a set of inputs should control an
entire app. Put the shared data logic in one reactive expression, then
let each tab focus only on its own view of that shared state.
