---
title: "Rhino: When a Shiny App Becomes a Software Project"
author: "Dror Berel"
format:
  html:
    page-layout: full
    toc: true
    toc-location: left
    toc-depth: 3
    self-contained: true
execute:
  echo: true
editor_options: 
  chunk_output_type: console
editor: 
  markdown: 
    wrap: 72
---

# Learning goals

By the end of this lesson, you should be able to:

- explain what Rhino adds on top of standard Shiny
- compare a regular Shiny app structure with a Rhino app structure
- recognize when Rhino is useful and when it is probably too much
- write a small Shiny example in both styles
- understand the basic Rhino workflow

# What is Rhino?

Rhino is an R package and project framework for building Shiny apps. It
does not replace Shiny. It gives Shiny apps a more formal project
structure.

Standard Shiny answers this question:

> How do I make an interactive web app in R?

Rhino answers a different question:

> How do I organize, test, and maintain a Shiny app as it grows?

That difference matters. A small app can be easier to build with plain
Shiny. A larger app with multiple developers, modules, tests, and
deployment requirements can benefit from Rhino.

# Packages

Install these packages if needed:

```{r}
# install.packages(c("shiny", "rhino", "ggplot2", "dplyr"))
```

For most of this lesson, the examples are shown as code snippets rather
than launched inside this Quarto document.

# Standard Shiny in one file

The smallest useful Shiny app can live in one `app.R` file.

```{r, eval=FALSE}
library(shiny)
library(ggplot2)

ui <- fluidPage(
  titlePanel("Standard Shiny"),
  sidebarLayout(
    sidebarPanel(
      selectInput(
        "x",
        "X variable",
        choices = c("disp", "hp", "wt", "qsec"),
        selected = "wt"
      )
    ),
    mainPanel(
      plotOutput("scatter")
    )
  )
)

server <- function(input, output, session) {
  output$scatter <- renderPlot({
    ggplot(mtcars, aes(x = .data[[input$x]], y = mpg)) +
      geom_point(size = 3) +
      labs(x = input$x, y = "mpg")
  })
}

shinyApp(ui, server)
```

This is a good structure when:

- the app is small
- one person owns most of the code
- the app has only a few inputs and outputs
- there are few reusable pieces
- testing is mostly manual

There is nothing wrong with this. Many useful Shiny apps should stay this
simple.

# Where standard Shiny starts to feel crowded

As a Shiny app grows, `app.R` often starts collecting many different
responsibilities:

- UI layout
- data loading
- data cleaning
- reactive logic
- plots
- tables
- modules
- helper functions
- CSS and JavaScript
- tests or deployment scripts

The app may still work, but it becomes harder to answer basic questions:

- Where should new code go?
- Which functions are safe to reuse?
- How do we test this piece without running the whole app?
- How do two people work on the app without editing the same file?
- How do we keep code style consistent?

Rhino is designed for this stage.

# The same idea as a module

Before thinking about Rhino, it helps to separate the app into a Shiny
module. This is still standard Shiny.

```{r, eval=FALSE}
library(shiny)
library(ggplot2)

scatter_ui <- function(id) {
  ns <- NS(id)

  tagList(
    selectInput(
      ns("x"),
      "X variable",
      choices = c("disp", "hp", "wt", "qsec"),
      selected = "wt"
    ),
    plotOutput(ns("scatter"))
  )
}

scatter_server <- function(id, data) {
  moduleServer(id, function(input, output, session) {
    output$scatter <- renderPlot({
      ggplot(data(), aes(x = .data[[input$x]], y = mpg)) +
        geom_point(size = 3) +
        labs(x = input$x, y = "mpg")
    })
  })
}

ui <- fluidPage(
  titlePanel("Standard Shiny with a module"),
  scatter_ui("cars")
)

server <- function(input, output, session) {
  cars <- reactive(mtcars)
  scatter_server("cars", cars)
}

shinyApp(ui, server)
```

This is already better. The plot feature has a clear UI function and a
clear server function.

Rhino builds on this kind of thinking. It encourages the app to be made
from small, named pieces instead of one long script.

# Creating a Rhino app

A Rhino app starts as a project.

```{r, eval=FALSE}
rhino::init("cars_app")
```

That creates a folder structure for the app. A typical Rhino project has
places for app code, tests, static assets, and configuration.

The exact folder contents can vary by Rhino version, but the important
idea is stable:

- app code lives in a predictable place
- modules are stored as separate files
- app startup is separated from feature code
- tests and tooling are part of the project from the beginning

# A small Rhino-style module

In a Rhino project, a feature usually lives in its own module file.

For example, a file like `app/view/scatter.R` might contain:

```{r, eval=FALSE}
box::use(
  ggplot2[ggplot, aes, geom_point, labs],
  shiny[NS, moduleServer, plotOutput, renderPlot, selectInput, tagList]
)

ui <- function(id) {
  ns <- NS(id)

  tagList(
    selectInput(
      ns("x"),
      "X variable",
      choices = c("disp", "hp", "wt", "qsec"),
      selected = "wt"
    ),
    plotOutput(ns("scatter"))
  )
}

server <- function(id, data) {
  moduleServer(id, function(input, output, session) {
    output$scatter <- renderPlot({
      ggplot(data(), aes(x = .data[[input$x]], y = mpg)) +
        geom_point(size = 3) +
        labs(x = input$x, y = "mpg")
    })
  })
}
```

This looks similar to a normal Shiny module, but there are two important
differences:

- the module is in its own file by default
- dependencies are imported explicitly with `box::use()`

# What does `box::use()` do here?

Rhino uses the `box` package to make R code behave more like modules in
Python or JavaScript.

In regular Shiny, we usually write:

```{r, eval=FALSE}
library(shiny)
library(ggplot2)
```

That loads the packages and attaches many exported names into the current
search path. After `library(shiny)`, functions like `fluidPage()`,
`selectInput()`, and `renderPlot()` are available by name.

That is convenient, but it is also broad. In a large app, it can become
harder to know where a function came from, and two packages can export
functions with the same name.

With `box::use()`, each file says exactly what it needs:

```{r, eval=FALSE}
box::use(
  ggplot2[ggplot, aes, geom_point, labs],
  shiny[NS, moduleServer, plotOutput, renderPlot, selectInput, tagList]
)
```

Read this as:

- from `ggplot2`, import only `ggplot()`, `aes()`, `geom_point()`, and
  `labs()`
- from `shiny`, import only `NS()`, `moduleServer()`, `plotOutput()`,
  `renderPlot()`, `selectInput()`, and `tagList()`

Those names are then available inside this file:

```{r, eval=FALSE}
plotOutput(ns("scatter"))

output$scatter <- renderPlot({
  ggplot(data(), aes(x = .data[[input$x]], y = mpg)) +
    geom_point(size = 3) +
    labs(x = input$x, y = "mpg")
})
```

This makes the dependency list part of the module. If this file uses
`selectInput()`, the import line tells us it came from `shiny`. If this
file uses `geom_point()`, the import line tells us it came from
`ggplot2`.

# Common `box::use()` patterns

There are a few useful ways to read and write `box::use()`.

Import selected functions:

```{r, eval=FALSE}
box::use(
  shiny[selectInput, plotOutput, renderPlot]
)
```

This makes those functions available directly as `selectInput()`,
`plotOutput()`, and `renderPlot()`.

Import a package without attaching its functions:

```{r, eval=FALSE}
box::use(
  ggplot2
)
```

Now the package is available as `ggplot2`, but its functions are not
attached directly. You would call them with `$`:

```{r, eval=FALSE}
ggplot2$ggplot(mtcars, ggplot2$aes(wt, mpg)) +
  ggplot2$geom_point()
```

This is very explicit, but it can feel verbose for common plotting code.
That is why the earlier example imports only the few `ggplot2` functions
it needs.

Import a package with an alias:

```{r, eval=FALSE}
box::use(
  plt = ggplot2
)

plt$ggplot(mtcars, plt$aes(wt, mpg)) +
  plt$geom_point()
```

This can be useful when a package name is long or when you want to make
the code read more clearly.

Import all exported functions:

```{r, eval=FALSE}
box::use(
  shiny[...]
)
```

The `...` means "all exported names from this package". This is closest
to `library(shiny)`. It is usually better to avoid this in Rhino modules
because the goal is to keep each file's dependencies clear.

# Importing local Rhino modules

The same `box::use()` syntax also imports local files from the Rhino
project.

This line imports a local module:

```{r, eval=FALSE}
box::use(
  app/view/scatter
)
```

It points to a file such as:

```default
app/view/scatter.R
```

After that import, the module is available as `scatter`. The functions
defined in that file are accessed with `$`:

```{r, eval=FALSE}
scatter$ui("cars")
scatter$server("cars", cars)
```

In this tutorial, `scatter.R` defines two functions:

- `ui()` creates the Shiny UI for the feature
- `server()` creates the Shiny server logic for the feature

So `scatter$ui()` means:

> use the `ui()` function from the `scatter` module

And `scatter$server()` means:

> use the `server()` function from the `scatter` module

This is the main reason `box` fits Rhino well. Each Shiny feature can be
a regular `.R` file, and the main app can import those feature files as
modules.

# Using the module in the app

The main app file imports the module and calls its UI and server pieces.

For example, a file like `app/main.R` might contain:

```{r, eval=FALSE}
box::use(
  shiny[fluidPage, moduleServer, reactive, titlePanel],
  app/view/scatter
)

ui <- function(id) {
  fluidPage(
    titlePanel("Rhino Shiny app"),
    scatter$ui("cars")
  )
}

server <- function(id) {
  moduleServer(id, function(input, output, session) {
    cars <- reactive(mtcars)
    scatter$server("cars", cars)
  })
}
```

The main app now reads more like a table of contents. It tells us which
features are used, but the feature code lives elsewhere.

# Standard Shiny versus Rhino

| Topic | Standard Shiny | Rhino |
|---|---|---|
| Best starting point | One `app.R` file | A project scaffold |
| App size | Small to medium | Medium to large |
| Learning curve | Lower | Higher |
| File structure | Flexible | Opinionated |
| Modules | Optional | Expected |
| Testing | Possible, but you decide how | Encouraged from the start |
| Team work | Can become messy | Easier to organize |
| Tooling | Add as needed | Built into the project workflow |

The tradeoff is simple:

- standard Shiny is faster for small apps
- Rhino is better when structure matters more than speed of setup

# When should you use Rhino?

Use Rhino when at least a few of these are true:

- the app will be maintained for months or years
- more than one developer will work on it
- the app has many modules or pages
- the app has important business logic
- the app needs automated tests
- the app will be deployed repeatedly
- you want consistent style, linting, and project conventions
- the app is closer to a product than a prototype

Rhino is especially useful for production Shiny apps. In that setting,
the app is not just a script. It is software that needs maintenance,
review, testing, and deployment.

# When should you stay with standard Shiny?

Stay with standard Shiny when:

- the app is a quick prototype
- the app is for a one-time analysis
- the app has only one or two screens
- the app is owned by one person
- the app logic is simple
- the team is still learning basic Shiny

For a small app, Rhino can feel like extra ceremony. That is not a Rhino
problem. It means the app does not need that much structure yet.

# A practical rule

Start with standard Shiny when you are exploring.

Move toward Rhino when the app becomes something people depend on.

One practical signal is when you start saying:

> I am afraid to change this file because I do not know what else it
> affects.

That is a good time to add stronger structure. Rhino gives you that
structure.

# Small exercise

Take this small standard Shiny app:

```{r, eval=FALSE}
library(shiny)

ui <- fluidPage(
  titlePanel("Iris summary"),
  selectInput("species", "Species", unique(iris$Species)),
  tableOutput("summary")
)

server <- function(input, output, session) {
  filtered <- reactive({
    iris[iris$Species == input$species, ]
  })

  output$summary <- renderTable({
    data.frame(
      rows = nrow(filtered()),
      mean_sepal_length = mean(filtered()$Sepal.Length),
      mean_petal_length = mean(filtered()$Petal.Length)
    )
  })
}

shinyApp(ui, server)
```

Practice questions:

- What code belongs to the feature?
- What code belongs to app startup?
- What would you put in a module file?
- What behavior would be worth testing?

A reasonable module boundary is the iris summary feature:

- `ui(id)` creates the species selector and summary table
- `server(id, data)` filters the data and renders the table
- a separate helper function could calculate the summary values

# Key takeaways

- Rhino is not a different kind of Shiny app. It is a structured way to
  build Shiny apps.
- Standard Shiny is best for small apps, learning, prototypes, and simple
  dashboards.
- Rhino is best for larger apps that need maintainability, tests, team
  development, and production discipline.
- Do not choose Rhino because it is more advanced. Choose it when the app
  needs the structure.
